---
title: "Lab: Week 4"
author: "36-350 -- Statistical Computing"
date: "Week 4 -- Spring 2022"
output:
  pdf_document:
    toc: no
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
---


```{r wrap-hook,echo=FALSE}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})
```

---

# Apply: Base R

You are given the following 8 $\times$ 8 matrix:
```{r linewidth=80}
set.seed(1001)
mat = matrix(rnorm(64),nrow=8)
mat[5,8] = mat[6,7] = mat[4,2] = NA
```

## Question 1

*Notes 4A (6,10,12)*

Compute the mean for each row and for each column using both `apply()` *and* either `rowMeans()` or `colMeans()`. (So there should be four function calls overall.) Deal with the `NA`s by passing (an) additional argument(s) to these functions, when possible.
```{r linewidth=80}
apply(mat, 1, mean, na.rm = TRUE)
apply(mat, 2, mean, na.rm = TRUE)
rowMeans(mat, na.rm = TRUE)
colMeans(mat, na.rm = TRUE)
```


## Question 2

*Function writing review*

How does the `Income` variable in `R`'s `state.x77` matrix correlate with other variables? Write a function called `cor_var()` that takes two inputs: `v1`, a numeric vector; and `v2`, another numeric vector whose default value is `state.x77[,"Income"]`. Its output should be the correlation of `v1` and `v2`, computed via the `cor()` function. Check that `cor_var(v1=state.x77[,"Life Exp"])` gives you 0.3402553, and `cor_var(v1=state.x77[,"Income"])` gives you 1.
```{r linewidth=80}
cor_var = function(v1, v2 = state.x77[,"Income"])
{
  cor(v1, v2)
}
cor_var(v1=state.x77[,"Life Exp"])
cor_var(v1=state.x77[,"Income"])
```

## Question 3

*Notes 4A (6-7,9-10)*

Using `apply()` and the function `cor_var()` that you defined in the last question, calculate the correlation between each one of the 8 variables in the `state.x77` matrix and the `Population` variable. Display these correlations.
```{r linewidth=80}
apply(state.x77, 2, cor_var, v2=state.x77[,"Population"])
```

## Question 4

*Notes 4A (6,10)*

Using `apply()` and the base `R` `stats` package function `cor()`, display the Spearman correlation between each one of the eight variables in the `state.x77` matrix and the `Frost` variable. (Note that `Spearman` is not the default value for the `method` argument to the `cor()` function.)
```{r linewidth=80}
apply(state.x77, 2, FUN=cor, y=state.x77[,"Frost"], method='spearman')
```

# Variations on Apply: Base R

## Question 5

*Notes 4B (6)*

Create a data frame called `state.df` from the matrix `state.x77` and the factors `state.region` and `state.division`. Be sure to name the two new columns appropriately. Using `state.df` and `tapply()`, compute the average population in each of the four defined regions of the U.S. Display the name of the region has the largest average population (and only that name). Then compute the average population in each of the nine defined divisions of the U.S., and display the name of the division has the largest average population (and only that name). Hint: the names may be displayed using a combination of `names()` and `which.max()`.
```{r linewidth=80}
# FILL ME IN
```

## Question 6

*Notes 4A (5) and Notes 4B (3-4)*

Split the rows of the data frame `state.df` by state divisions, and call the resulting list `state.df.by.div`. Then use `lapply()` to display just the first two rows of each data frame in the list `state.df.by.div`.
```{r linewidth=80}
#
```

---

Below, we read in a data table showing the fastest women's 100-meter sprint times.
```{r linewidth=80}
sprint.df = read.table("http://www.stat.cmu.edu/~mfarag/350/women_100m_with_header.dat",
                       header=TRUE,stringsAsFactors=FALSE)
class(sprint.df)
head(sprint.df)
```

---

## Question 7

*Review of string processing*

Extract the last four digits of each entry of the `Date` column. (Hint: you will have to use `as.character()` to convert `sprint.df$Date` from a factor variable to strings.) Create a new data frame called `new.sprint.df` that combines `sprint.df` and a new column called `Year` that contains your extracted four-digit years. Display the first five rows and all 11 columns of `new.sprint.df`. Display the class of the newly created `Year` column.
```{r linewidth=80}
# FILL ME IN
```

## Question 8

*Notes 4B (6)*

Using `tapply()` and the newly created `Year` column, compute the median 100-meter sprint time in each year of the data frame. Call the resulting vector `med.time.by.year`. Create a table of median times. Which median time appears the most, and how many times does it appear? When is the last year that that particular median time appeared in the data?
```{r linewidth=80}
# FILL ME IN
```
```
FILL ME IN
```

---

Below, we read in a data table related to the political economy of strikes.
```{r linewidth=80}
strikes.df = read.csv("http://www.stat.cmu.edu/~mfarag/350/strikes.csv")
class(strikes.df)
head(strikes.df)
dim(strikes.df) # Note that since 18 × 35 = 630 > 625, some years missing from some countries
```

---

## Question 9

*Notes 4A (5) and Notes 4B (5)*

Split `strikes.df` by country, using the `split()` function. Call the resulting list `strikes.by.country`. Using `strikes.by.country` and `sapply()`, compute the average centralization metric (a quantity related to unionization) for each country over the range of years in the file. Display the names of the countries that had the highest and lowest average centralization metric (and only the names of those countries).
```{r linewidth=80}
# FILL ME IN
```

## Question 10

*Notes 4B (5)*

Using `strikes.by.country` and `sapply()`, compute a summary of the long-term centralization metric for each country. Study the output---do its dimensions make sense to you?
```{r linewidth=80}
# FILL ME IN
```
```
FILL ME IN
```

## Question 11

*Notes 4B (5)*

Using `strikes.by.country` and just *one* call to `sapply()`, compute the average unemployment rate, average inflation rate, and average strike volume for each country. The output should be a matrix of dimension 3 x 18. Also, within that call, give the output matrix appropriate row names.
```{r linewidth=80}
# FILL ME IN
```

## Question 12

*Notes 4B (5)*

Using `strikes.df`, `split()`, and `sapply()`, compute the average unemployment rate for each country, before and during 1970, and after 1970. Display the output; it should be a numeric vector of length 36. One way to perform the splitting is to define a new column called `pre1970` that indicates that a year column is less than or equal to 1970. Then use both `country` *and* `pre1970` to do the splitting. If you are not sure how to use both factor variables at once, look at the documentation for `split()`, specifically its argument `f`.
```{r linewidth=80}
# FILL ME IN
```

## Question 13

*Review of matrices*

Using the result from above, display the difference in the average unemployment rate before and after 1970 for each country. (To be clear: subtract the pre-1970 results from the post-1970 results.) Which country had the biggest increase in average unemployment from before to after? The biggest decrease? (Hint: use the output from Q12 to populate a matrix, with pre-1970 results in one column and post-1970 results in another.)
```{r linewidth=80}
# FILL ME IN
```

---

## Question 14

*Pipes + Notes 4D (6)*

How does the `Frost` variable in `R`'s `state.x77` matrix correlate with other variables? Cast `state.x77` to a data frame, and, using pipes, generate the correlation matrix for `Frost` and `Life.Exp`. (Note that the act of casting changed the name of the life expectancy column from `Life Exp` to `Life.Exp`.) The off-diagonal elements of the matrix should be 0.262068.
```{r linewidth=80}
# FILL ME IN
```

## Question 15

*Pipes + Notes 4D (9)*

Take the `state.df` data frame defined in Q5 above and mutate it so as to create a new column: `GradLit`. This column should have, for each row in the data frame, the percentage of high school graduates divided by the percentage of literate (note: *literate*, not *illiterate*) individuals, times 100. Then pipe the output so as to compute the median value of `GradLit`. (There is a bit of weirdness here: due to environmental issues, your call to `median()` will not work unless it is placed within curly braces. You are only surrounding `median()` with curly braces...not the entire pipe stream!) Your final value should be 53.59844.
```{r linewidth=80}
# FILL ME IN
```

## Question 16

*Pipes + Notes 4D (5,6,8)*

Take the `state.df` data frame and (1) select all states in the South region, and (2) display the result ordered by the decreasing product of income and life expectancy. In the end, display just the state name and the computed product. There is a quirk here: selecting rows can lead to the loss of row names. (This means that here, you will have a final result but not know which states they correspond to.) To preserve the identity of the states, pipe `state.df` to the function `rownames_to_column("give column name here, like State")`, then do the rest of your piping.
```{r linewidth=80}
# FILL ME IN
```
